|

### 2022-02-02 text/ManuscriptDraft.Rmd
Manuscript update
Added figures and tables. Added examples of annotated images.

---
new_session: FALSE
title: |
  **Towards individual-based pollination ecology: Automatic tracking of life histories of individual flowers**
author:
- Hjalte M. R. Mann
- Alexandros Iosifidis
- Toke T. Høye
date: "`r format(Sys.time(), '%B %d, %Y')`"
output:
  word_document: default
  pdf_document:
    fig_caption: yes
    keep_tex: yes
    latex_engine: pdflatex
  html_document:
    df_print: paged
fontfamily: mathpazo
fontsize: 12pt
geometry: margin = 1in
header-includes:
- \usepackage{setspace}\doublespacing
- \usepackage[left]{lineno}
- \linenumbers
- \usepackage{dcolumn}
- \usepackage{caption}
- \usepackage{float}
- \usepackage{afterpage}
- \usepackage{siunitx}
- \usepackage{amsmath}
keywords: Tracking, flowering phenology, Arctic, DBSCAN
bibliography: ./library.bib
csl: ./journal-of-ecology.csl
abstract: ABSTRACT | A fundamental problem in pollination ecology is that plant reproduction consequences of pollinator visits to a flower throughout its lifespan is very difficult to track. Often simple statistics are used to describe the flowering phenology of a population of plants, e.g. onset or peak of flowering. Similarly, flower visitation rates are based on crude observation of flower populations. Here we show that image-based monitoring of field plots at very high temporal resolution can return information on flowering phenology at the level of individuals. Further, we present a framework for automatically tracking, filtering, and evaluating flowers in time-lapse image series. We compare the results of automatic tracking with manual tracking. Individual flower tracking could facilitate finegrained studies on plant reproductive success.... and we show that performance achieved by following automatic tracking can be reliable enough to estimate X, Y, Z...
---

@@

@@

Phenology is often studied with first events as proxies for the full phenology distribution of the population . For flowering plants the time of first flowering is used to represent the phenology of a population of plants. The ideal metric for phenology depends on the topic of the question, but often, the choice of onset of flowering as the proxy for phenology is based on data availability and not because it is necessarily considered the ideal metric. First events, such as onset of flowering, may in some cases be particularly ill suited for estimating sensitivity to climate change [@miller2008]. The date for first flowering represents the very extreme of the phenology distribution of a population and may be confounded with e.g. the size of the population considered. Further, for many questions the phenology distribution of a population is required, for example to quantify resource availability through the season. However, even detailed information on phenology at the level of a population is still an aggregation of the phenological distributions of the individual members of the population for which important variation may be masked [@inouye2019]. Having at hand the full distribution of flowering phenology at the level of individuals allows to extract relevant characteristics of the distribution and test their sensitivity towards environmental factors, but acquiring such data requires efficient and preferably automatized methods.

@@

Interactions require abundance of individuals in life-history stages in which interactions occur. Thus, interaction strength depends on overlap in phenology between individuals in these stages. Ideally, investigation of the sensitivity of interaction strength to dynamics in phenology requires quantification of phenology through the season for the species in question in order to asses both their phenology and abundance [@inouye2019; @linden2018].

If interacting and interdependent species, e.g. plants and pollinators, react to different environmental cues, or react differently to the same cues, their phenology may be desynchronized and temporal mismatches occur. Changes in flower phenology as a consequence of climate change may indirectly affect the abundance of flower-dependent species [@hoye2013] and these effects can be stronger than the direct effects of climate change [@ogilvie2017]. However, the relation between flower phenology and reproductive success is multifaceted and not simply a question of temporal overlap with pollinators. For example, @bolmgren2015 found that for an insect-pollinated plant species early flowering was associated with increased fruit-set but late flowering was associated with higher seed mass. Further, periods of resource scarcity through the season may be more important than the floral sum of the season, i.e. dynamics in resource availability above a level of sufficiency may have a weak effect on pollinator abundance while the time above this threshold has a stronger effect [@ogilvie2017].

Different phenological factors, such as onset, peak, and end of flowering may respond differently to climate change [@cook2012] and estimation of the degree of mismatch between species may thus be sensitive to the choice of metric [@inouye2019]. Efficient and precise methods for deriving full seasonal distributions of phenology at the level of individuals is much needed but so far most often unfeasible for logistical reasons.

Monitoring flower phenology at high temporal resolution is time-consuming, particularly in logistically challenging environments such as the Arctic. Consequently, flowering phenology of a population is often characterized simply by the date for onset of flowering, typically derived from weekly observations of sample plots [@Prevey2021]. Such inaccurate estimates may fail to reveal important dynamics in flowering phenology through the season. Automatic image-based monitoring of flowering phenology can return phenology data for specific species at very high temporal resolution (Mann et al., in review), but phenological responses at the individual level may be indiscernible regardless of the temporal resolution of the data at population level. For example, a shortening of individual flower longevity may not be directly obvious at the population level. In fact, many research questions can only be explored on the basis of individual phenology data. For example, investigating the association between reproductive success and timing of flowering or flower longevity requires phenology data at the level of individuals. Similarly so for investigating whether flower visitation rates and/or reproductive success depends on the timing of flowering for the individual flower.

A lot remains to be discovered about when and how the dynamics of pollination affects the reproductive success of flowering plants. High frequency observations of flower visits coupled with information on phenology and reproductive success at the level of individuals is required to unravel the relationship between pollinator visitation and plant reproductive success and the sensitivity of this interplay to climate.

In this paper, we show that information on phenology at the level of individuals can be derived from image-based monitoring of flower phenology. We present and evaluate an automatic flower tracking and filtering algorithm. As the data extracted from time-lapse series of flower plots through season violate the assumptions of standard tracking mechanism, we base the track assignment on the minimum distance between centroids of bounding boxes and introduce three user defined parameters that can help optimise tracking performance.

@@

Our method for tracking individual flowers facilitates the possibility of assigning indicators of reproductive success and factors driving variation in seed set to the individual. For instance, it could be explored whether seed set is affected by timing and length of flowering and/or abiotic conditions such as frost events. Hereby, the longstanding question of what drives variation in reproductive success of individual flowers could be explored. By simultaneously tracking flower visits, these could be assigned to the individual flower and visitation rates per flower could be calculated per flower and related to reproductive success. Further, any information of taxonomic grouping could facilitate comparative studies on pollination effectiveness between taxonomic groups of pollinators. The possibility of automatically tracking phenology at the level of individuals will facilitate investigation of a set of yet untestable research questions and is an important contribution to the shift towards efficient and detailed monitoring of ecological processes.

# Material and methods
#
## Study site and species
This study was based on five time-lapse image-series collected at the following three Arctic sites (sampling years given in brackets): Narsarsuaq, South Greenland (2018, 2019), Thule, North-West Greenland (2018), Ny-Ålesund, Svalbard (2019). The images were collected using time-lapse cameras (Moultrie Wingscapes TimelapseCam Pro, Moultrie Products, Birmingham Alabama, USA) permanently positioned approx. 60 cm above areas with either Dryas integrifolia in Narsarsuaq and Thule or D. octopetala in Ny-Ålesund. The white flowers of these two species are insect-pollinated and held erect above cushions of foliage. The species are native to and widespread in Asian, North American, and European arctic and alpine regions. They are geographically separated apart from a possible hybrid zone in Northeast Greenland [@philipp2003].

#
## The image series and flower annotations
Images were collected throughout the full growing season. While the original time-lapse frequency was higher, the basis for this study was 1-hr interval sereis. We limited the 1-hr interval image-series to the flowering season and randomly randomly sampled images from each series. The number of images for each series after sampling is given in table 1. All flowers in each of these images were manually annotated using the rectangular bounding box tool in the VIA VGG annotation software [@dutta2019] and each individual flower was assigned a unique ID. These annotations constitute our ground truth tracks. The series varied substantially in the number of flowers they contained (table 1). Figure 1 show annotated example images from three series (xxx) at different times during the flowering season (TO-DO).

@@

#
![](../figures/SOmeFigure.png){ width=100% }

**Figure 1:** Figure text... Examples of annotated images (TO-DO)

@@

Several issues complicate the task of tracking individual flowers through time-lapse image-series.

First, stand-alone monitoring of the relatively slow process of flower development through full growing seasons (several months) requires only low frame rates. Here we used several images per day. These frame rates capture the phenology of the flowers at very high detail, but fail to capture the full movement of flowers. For example, as the wind shifts, the flower heads change direction, which in the image series happens instantaneously (i.e. between two consecutive frames). However, as they are constrained by their stalk, there is a limit to the distance they can move. With the tracking parameter **running mean**, we have the option to base the tracking on the distance between a point in the current frame and the running mean of the positions of a number of previous points in a track, taking advantage of the fact that flower move around a central point.

Second, as flowers move, they may temporarily occlude the view of other flowers and flowers close to the edge of the frame may move in and out of view. This can cause a track to be lost and a new track erroneously being initiated when the flower reappears. However, if a flower reappears in the same area as a flower is already being tracked after disappearing in a few frames, it is reasonable to assume that it is the same individual and not that the old flower wilted/disappeared and a new one developed. The parameter **max gap size** sets the number of frames a track can be lost before a new track is initiated for points appearing in the same area. If a given flower has not been detected in a few frames, the track will not be lost. Concurrently, this deals with potential false negatives. The counter for number of disappeared frames for a track is reset when a new point is associated with the track within the threshold.

Third, a centroid tracking algorithm will associate a point with a tracking based on minimum distance only, disregarding the absolute distance. As flowers are constrained in the movement, we can assume a maximum distance that will occur between points and force the initiation of tracks for points that exceed this threshold. We set this threshold with the parameter **max distance**.

# EDIT FIGURES: Zoom in, bigger dots, combine into one figure.
#
![](../figures/figure_1.png){ width=100% }

**Figure 1:** `r word_comment("The three figures below are quick mockups to explain the challenges with tracking flowers. Not sure if figures are needed or text is sufficient.", "Simple")` centroid tracking may produce erroneous associations when objects move between frames. Blue shows detections in the current frame (bounding box and centroid point). Red shows centroid points for the detections in the previous frame.

# 
![](../figures/figure_22.jpg){width=100%}

**Figure 2:** Simple centroid tracking may produce erroneous associations when objects move between frames. Basing the association on the running mean of the positions of the previous n number of tracks may alleviate this issue. In this case basing the association on only the previous point would produce a wrong results while basing it on the running mean would produce a correct result. Red points: Centroids for bounding boxes in current frame, blue points: Centroids for bounding boxes in previous frame; grey points: Centroids for bounding boxes in a number of frames before t-1; green points: Running mean of the previous n points. Circles delimit the two individuals.

# 
![](../figures/figure_3.png){ width=100% }

**Figure 3:** Simple centroid tracking may produce erroneous associations when objects disappear periodically from the frame. Here the top flower moves out of frame and the bottom flower would be assigned to the track of the top flower in frame t-1.

@@

We base our filtering on a density-based clustering of the centroid of a geometry of each track. The geometries are derived in the following way: For tracks consisting of only a single point, the coordinates of the point are used as the centroid. For tracks consisting of two points, we establish the straight line between the points and calculate the centroid of the line. For tracks consisting of three points, we establish the triangle from the points and calculate the centroid of the triangle. For tracks consisting of four points or more, we calculate the convex hull of all the points included in the track and derive the polygon from the vertices of the convex hull. Hereafter we calculate the centroid of the polygon. For tracks that contain more than two points but where the points a collinear, we establish the line through the points and calculate the centroid of the line.

The filtering is done on the geometry centroids to remove tracks in areas with a high density of tracks as these have a high risk of tracking mismatches. The filtering is done in two steps. First, the DBSCAN clustering algorithm [@ester1996] is run on the centroids with a conservatively high value for the eps parameter, meaning that tracks in close vicinity to each other will be clustered together. Second, all tracks that were not assigned to a unique cluster are removed. Tracks that are spatially isolated remains.

We demonstrate the filtering method on the three series for which our tracking algorithm did not return perfect results. For this, we chose a fixed combination of the tracking algorithm parameters across all three series (max gap size: 10; running mean: 10; max distance: 300).

We then applied the filtering method on the three series with a range of values for eps (x-xx, step size of 10) and identified the value that returned the highest sum of tracks across the three series at the lowest number of mismatches. We present the tracking accuracy before filtering and on the remaining tracks after filtering. We underline that the results presented are thus highly conservative. I.e., the approach could be optimised to extract a higher number of tracks while maintaining the number of mismatches. For example, basing the filtering on the best performing tracking parameters for each series specifically would improve the output of the tracking algorithm. Similarly, the results could be improved by fine-tuning the value for eps for each series individually. However, as our goal here was to demonstrate that our tracking and filtering method can be applied in a naive setting with fixed values for each parameter we have not included those results.

@@

@@

The results of the tracking parameter test are given in table 1. We present the tracking performance without the use of the three tracking parameters along with the best performance (lowest number of mismatches/highest MOTA), and the best performance where the number of tracks correspond to the number of individual flowers in the series. When several values for a parameter was included in a result, we present the range of values. For two series, 2018 THUL B and 2019 NARS D, the tracking algorithm returned perfect results for several combinations of parameters. For the three remaining series, the algorithm returned consistently high MOTA scores (above 0.95). In all cases, except 2019 NARS D, where the tracking algorithm returned perfect results with all parameters zeroed, did the inclusion of the user set tracking parameters improve the tracking accuracy.

In one case, NARS-13, four different numbers of tracks were returned among the parameter combinations that resulted in the lowest number of mismatches. A number of tracks equal to the number of flowers in the series were among these four.

In one case, NYAA-04, the parameter combination that produced the highest accuracy did not return a number of tracks equal to the number of flowers in the series. However, combinations that did produce this number of tracks also returned a higher number of mismatches.

**Table 1:** Table description... To-do: Rename series to camera codes.

#
![](../figures/table1.png){ width=100% }

@@

The results of the filtering algorithm on the three series in which our tracking algorithm did not return perfect results are given in table 2 and the filtering process is visualised in fig. 4. Our filtering method extracted 28 tracked flowers from the three series with complex scenes using a fixed value of 350 for eps and fixed values for the tracking parameters (10, 10, 300, for running mean, max gap size, and max distance, respectively) with only a single mismatch included in the 28 tracks.

**Table 2:** Results of the filtering algorithm with an eps value of 350 applied across all three series To-do: Change image series to camera codes.

![](../figures/FilterScores.png){ width=100% }

@@

**Figure 4:** Track filtering process. Row 1, 2, and 3 is NARS-13, NARS-04, and NYAA-04, respectively. First column shows the centroid points in the given series, coloured by the track id from the centroid tracking algorithm. Second column shows the polygons calculated from the tracks. The DBSCAN clustering algorithm with eps = 350 was applied to the centroids of these polygons and the polygons are coloured by cluster id. Third column shows the results of the filtering where all tracks that were not assigned a unique cluster has been removed.

# Discussion

#
## Tracking

Many tracking algorithms, such as the Simple online and real-time tracking (SORT) [@sort], use Kalman filtering to predict the future location of an object by estimating its velocity [@kalmann]. However, Kalman filtering assumes linear motion of the object and does not handle abrupt motion well. As the flowers are constrained in movement by their stalk, their movement does not follow the assumption of linear motion. Some tracking algorithms, such as DeepSORT [@deepsort2017], take the tracking-by-detection approach in which convolutional neural networks are applied to leverage appearance information in order to continuously distinguish individuals through time, thus aiding the tracking algorithm. In our case, however, flowers are not clearly distinguishable and one flower may look less alike between two frames than two flowers in a single frame.

Finally, some tracking methods require manual initiation of a fixed number of tracks for objects in the first frame and the algorithm will track these objects in the subsequent frames [@hu2012]. These methods does not allow new objects to enter the frame through the series. In our case, objects can disappear permanently from the frame, for example when flowers wilt, in which case the track should be terminated. Oppositely, new objects can continuously appear through the series when new flowers bloom of previously unseen flowers enter the frame. Additionally, objects may temporarily disappear from view and may or may not reappear at some later stage. These settings vastly complicate the task of tracking multiple objects.

Our tracking algorithm returned consistently accuracy despite the difficult settings.

@@

In a naive setting, a priori knowledge about which settings to use is not available. As with any tracking method, accuracy can only be assessed based on visual observation or more thoroughly by testing on subsample of tracks for which a ground truth exists. When applying the method in a new setting, we recommend performing a parameter test as demonstrated here on a subset of ground truth tracks to derive the optimal set of parameters.

@@

The two series for which the tracking algorithm produced perfect results were also the series with the lowest number of flowers. As the number of flowers increase so does the complexity of the task of tracking. A higher number of flowers within the frame increases the possibility of flowers in close vicinity to each other and of flowers overlapping and moving in and out of frame translating to a higher risk of making tracking errors. The very high complexity in the scenes with many flowers, especially NARS-13 with 85 flowers, renders it unlikely that any tracking method could produce perfect results. Although this setting can be considered an extreme case for our application, it underlines the relevance of a pipeline including both tracking and filtering.

@@

We presented a filtering method based on the density of tracks within the frame which can be applied subsequent of tracking. Extracting tracks that are spatially isolated does not guarantee that the tracks are correct/without errors. However, as spatially isolated objects are easier to track, it increases confidence in the remaining tracks.

>>>>

---
new_session: FALSE
#title: |
#  **Towards individual-based pollination ecology: Automatic tracking of life histories of individual flowers**
#date: "`r format(Sys.time(), '%B %d, %Y')`"
output:
  word_document: default
  pdf_document:
    fig_caption: yes
    keep_tex: yes
    latex_engine: pdflatex
  html_document:
    df_print: paged
fontfamily: mathpazo
fontsize: 12pt
geometry: margin = 1in
header-includes:
- \usepackage{setspace}\doublespacing
- \usepackage[left]{lineno}
- \linenumbers
- \usepackage{dcolumn}
- \usepackage{caption}
- \usepackage{float}
- \usepackage{afterpage}
- \usepackage{siunitx}
- \usepackage{amsmath}
#keywords: Tracking, flowering phenology, Arctic
bibliography: ./library.bib
csl: ./journal-of-ecology.csl
---

@@

Title:

**Towards individual-based pollination ecology: Automatic tracking of life histories of individual flowers**

#
Authors:

Hjalte M. R. Mann^1,2^, Alexandros Iosifidis^2^, Toke T. Høye^1^

#
Author affiliations:

1: Department of Ecoscience and Arctic Research Centre, Aarhus University, C.F. Møllers Allé 4-8, 8000 Aarhus C, Denmark  
2: Department of Electrical and Computer Engineering – Signal Processing and Machine Learning, Aarhus University, Finlandsgade 22, 8200 Aarhus N, Denmark

#
Corresponding author contact information:

Hjalte Mads Rosenstand Mann  
E-mail: mann@ecos.au.dk  
Phone: +45 31178585

#
ORCID IDs:

HMRM: 0000-0002-4768-4767, AI: 0000-0003-4807-1345, TTH: 0000-0001-5387-3284

#
Keywords:

Tracking, flowering phenology, Arctic, dryas octopetala, dryas integrifolia, ecological monitoring, life-history variation

\newpage
# Abstract

A fundamental problem in pollination ecology is that plant reproductive consequences of pollinator visits to a flower throughout its lifespan is very difficult to track. Deciphering the consequences of e.g. shifts in flower visitation rates to plant reproductive success and to identify mismatches between interacting and interdependent species such as insect-pollinated plants and pollinators throughout the season requires full-season phenology data, preferably at the level of individuals. Often simple statistics are used to describe the flowering phenology of a population of plants, e.g. onset or peak of flowering, which may fail to reveal the true dynamics in flowering and reproduction. Here we show that image-based monitoring of field plots at very high temporal resolution can return information on flowering phenology at the level of individuals. Further, we present a framework for automatically tracking, filtering, and evaluating flowers in time-lapse image series. We compare the results of automatic tracking with manual tracking. We show that performance achieved by automatic flower tracking can be reliable in obtaining individual flower phenology. Our method could facilitate fine-grained studies on plant reproductive success and is an important tool for estimating the sensitivity of plant-pollinator interactions to climate change. Supporting this paper is a publicly available web application through which users can apply the tracking and filtering methods on their own data.

@@

@@

Interactions require abundance of individuals in life-history stages in which interactions occur. Thus, interaction strength depends on overlap in phenology between individuals in these stages. Ideally, investigation of the sensitivity of interaction strength to dynamics in phenology requires quantification of phenology through the season for the species in question in order to asses both their phenology and abundance [@inouye2019; @linden2018]. If interacting and interdependent species, e.g. plants and pollinators, react to different environmental cues, or react differently to the same cues, their phenology may be desynchronized and temporal mismatches occur. Changes in flower phenology as a consequence of climate change may indirectly affect the abundance of flower-dependent species [@hoye2013] and these effects can be stronger than the direct effects of climate change [@ogilvie2017]. However, the relation between flower phenology and reproductive success is multifaceted and not simply a question of temporal overlap with pollinators. For example, @bolmgren2015 found that for an insect-pollinated plant species early flowering was associated with increased fruit-set but late flowering was associated with higher seed mass. Further, periods of resource scarcity through the season may be more important than the floral sum of the season, i.e. dynamics in resource availability above a level of sufficiency may have a weak effect on pollinator abundance while the time above this threshold has a stronger effect [@ogilvie2017].

Monitoring flower phenology at high temporal resolution is time-consuming, particularly in logistically challenging environments such as the Arctic. Consequently, flowering phenology of a population is often characterized simply by the date for onset of flowering, typically derived from weekly observations of sample plots [@Prevey2021]. Such inaccurate estimates may fail to reveal important dynamics in flowering phenology through the season. For example, different phenological factors, such as onset, peak, and end of flowering may respond differently to climate change [@cook2012] and estimation of the degree of mismatch between species may thus be sensitive to the choice of metric [@inouye2019]. A lot remains to be discovered about when and how the dynamics of pollination affects the reproductive success of flowering plants. High frequency observations of flower visits coupled with information on phenology and reproductive success at the level of individuals is required to unravel the relationship between pollinator visitation and plant reproductive success and the sensitivity of this interplay to climate.

The ideal metric for phenology depends on the topic of the question, but often the choice of onset of flowering as the variable representing phenology is based on data availability and not because it is considered the ideal metric. First events, such as onset of flowering, may in some cases be particularly ill suited for estimating sensitivity to climate change [@miller2008]. Onset of flowering represents the very extreme of the phenological distribution of a population and may be confounded with e.g. the size of the population considered. Further, for many questions, the phenology distribution of a population is required, for example to quantify resource availability through the season. However, even detailed information on phenology at the level of a population is still an aggregation of the phenological distributions of the individual members of the population for which important variation may be masked [@inouye2019]. Having at hand the full distribution of flowering phenology at the level of individuals allows to extract relevant characteristics of the distribution and test their individual sensitivities towards environmental factors, but acquiring such data requires efficient and preferably automatized methods. Automatic image-based monitoring of flowering phenology can return phenology data for specific species at very high temporal resolution (Mann et al., in review (Note: I'll update this if the paper gets accepted or remove it if not)), but phenological responses at the individual level may be indiscernible regardless of the temporal resolution of the data at population level. For example, a shortening of individual flower longevity may not be directly obvious at the population level. In fact, many research questions can only be explored on the basis of individual phenology data. For example, investigating the association between reproductive success and timing of flowering or flower longevity requires phenology data at the level of individuals. Similarly so for investigating whether flower visitation rates and/or reproductive success depends on the timing of flowering for the individual flower. Efficient and precise methods for deriving full seasonal distributions of phenology at the level of individuals is much needed.

Cameras have become an ubiquitous tool for non-invasive and efficient monitoring of natural processes [@howe2017; @glover2019], but the implementation of image-based collection of data has outpaced the development of reliable and easily implementable methods for processing and analysing the big data output from this transformation [@weinstein2018; @farley2018]. Computer vision and deep learning tools and being developed and applied for automatic processing and extraction of relevant data from these very large image sets [@willi2019;  @tabak2019] and increasingly for challenging tasks in entomology [@hoye2021], such as for fine-grained species classification [@spiesman2021] and monitoring of pollination events by detecting, identifying, and counting insects [@bjerge2021]. Collectively, the development and adoption of modern technology allows for collection and processing of ecological data at unprecedented scales.

Here, we use stand-alone time-lapse cameras to monitor the phenology of two Arctic flowering plant species Dryas integrifolia and D. octopetala. We show that information on phenology at the level of individuals can be derived from image-based monitoring of flower phenology. We present and evaluate an automatic flower tracking and filtering algorithm. As the data extracted from time-lapse series of flower plots through season violate the assumptions of standard tracking mechanism, we base the track assignment on the minimum distance between centroids of bounding boxes and introduce three user defined parameters that can help optimise tracking performance.

@@

Our method for tracking individual flowers facilitates the possibility of assigning indicators of reproductive success and factors driving variation in seed set to the individual. For instance, it could be explored whether seed set is affected by timing and length of flowering and/or abiotic conditions such as frost events. Hereby, the long-standing question of what drives variation in reproductive success of individual flowers could be explored. By simultaneously tracking flower visits, these could be assigned to the individual flower and visitation rates per flower could be calculated per flower and related to reproductive success. Further, any information of taxonomic grouping could facilitate comparative studies on pollination effectiveness between taxonomic groups of pollinators. The possibility of automatically tracking phenology at the level of individuals will facilitate investigation of a set of yet untestable research questions and is an important contribution to the shift towards efficient and detailed monitoring of phenology.

# Material and methods
## Study site and species
This study was based on five time-lapse image-series collected at the following three Arctic sites (sampling years given in brackets): Narsarsuaq (NARS), South Greenland (2018, 2019), Thule (THUL), North-West Greenland (2018), Ny-Ålesund (NYAA), Svalbard (2019). The images were collected using time-lapse cameras (Moultrie Wingscapes TimelapseCam Pro, Moultrie Products, Birmingham Alabama, USA) permanently positioned approx. 60 cm above areas with either *Dryas integrifolia* in Narsarsuaq and Thule or *D. octopetala* in Ny-Ålesund. The white flowers of these two species are insect-pollinated and held erect above cushions of foliage. The species are native to and widespread in Asian, North American, and European Arctic and alpine regions. They are geographically separated apart from a possible hybrid zone in North-East Greenland [@philipp2003].

#
## The image series and flower annotations
Images were collected throughout the full growing seasons. While the original time-lapse frequency was higher, the basis for this study was 1-hr interval series. We limited the 1-hr image-series to the include only the flowering season and randomly randomly sampled images from each series. The number of images for each series after sampling is given in table 1. All flowers in bloom in each of these images were manually annotated using the rectangular bounding box tool in the VIA VGG annotation software [@dutta2019] and each individual flower was assigned a unique ID. These annotations constitute our ground truth tracks. The series varied substantially in the number of flowers they contained (table 1). Examples of annotated images from the 2019 NARS C series at three different times in the flowering season are given in fig. 1. Figure 2 shows the time of flowering for each individual flower and the number of days the flower was present in the images.

@@

#
![](../figures/AnnotationExamples_3.png){ width=100% }

**Figure 1:** Examples of annotated images for 2019 NARS C. a: Onset of flowering at DOY 145; b: Peak flowering (52 annotated flowers) 6 days after onset (DOY 151); c: 6 days after peak flowering (DOY 157). Flower bounding boxes are colour coded by flower ID.

#
![](../figures/OnsetVSLength.png){ width=100% }  
**Figure 2:** Onset of flowering (DOY) and flower lifespan (days) for each individual flower in each of the five image-series.

Several issues complicate the task of tracking individual flowers through time-lapse image-series. First, stand-alone monitoring of the relatively slow process of flower development through full growing seasons (several months) requires only low frame rates. Here we used several images per day. These frame rates capture the phenology of the flowers at very high detail, but fail to capture the full movement of flowers. For example, as the wind shifts, the flower heads change direction, which in the image series happens instantaneously (i.e. between two consecutive frames) (fig. 3a). However, as they are constrained by their stalk, there is a limit to the distance they can move. With the tracking parameter **running mean**, we have the option to base the tracking on the distance between a point in the current frame and the running mean of the positions of a number of previous points in a track, taking advantage of the fact that flower move around a central point. Increasing the value for this parameter decreases the weighting of movement between individual frames.

Second, as flowers move, they may temporarily occlude the view of other flowers and flowers close to the edge of the frame may move in and out of view. This can cause a track to be lost and a new track erroneously being initiated when the flower reappears (fig. 3b). However, if a flower reappears in the same area as a flower is already being tracked after disappearing in a few frames, it is reasonable to assume that it is the same individual and not that the old flower wilted/disappeared and a new one developed. The parameter **max gap size** sets the number of frames a track can be lost before a new track is initiated for points appearing in the same area. If a given flower has not been detected in a few frames, the track will not be lost. Concurrently, this deals with potential false negatives. The counter for number of disappeared frames for a track is reset when a new point is associated with the track within the threshold.

Third, a standard centroid tracking algorithm will associate a point with a tracking based on minimum distance only, disregarding the absolute distance. As flowers are constrained in the movement, we can assume a maximum distance that will occur between points and force the initiation of tracks for points that exceed this threshold. We set this threshold with the parameter **max distance** measured as the euclidean distance in number of pixels.

#
![](../figures/HanddrawnFiguresIssues.png){ width=100% }

**Figure 3:** a: Centroid tracking may produce erroneous associations when objects move between frames. Here the track initiated on the top left flower in frame t-1 would switch to the bottom right flower in frame t b: Tracking errors may occur when objects temporarily disappear from the frame. Here the top flower moves out of frame and the bottom flower would be assigned to the track of the top flower in frame t-1 Red shows centroid of bounding box in the current frame and blue shows centroid points from the previous frame.

@@

@@

We base our filtering on the distance between geometry centroids the tracks. The geometries are derived in the following way: For tracks consisting of only a single point, the coordinates of the point are used as the centroid. For tracks consisting of two points, we establish the straight line between the points and calculate the centroid of the line. For tracks consisting of three points, we establish the triangle from the points and calculate the centroid of the triangle. For tracks consisting of four points or more, we calculate the convex hull of all the points included in the track and derive the polygon from the vertices of the convex hull. Hereafter we calculate the centroid of the polygon. For tracks that contain more than two points but where the points a collinear, we establish the line through the points and calculate the centroid of the line.

The filtering is done on the geometry centroids to remove tracks in close vicinity to each other as these have a high risk of tracking mismatches. The distances between all centroid tracks are calculated (euclidean distance in number of pixels) and only tracks that have no neighbouring centroid within the filtering distance threshold are kept.

We demonstrate the filtering method on the three series for which our tracking algorithm did not return perfect results. For this, we chose a fixed combination of the tracking algorithm parameters across all three series (max gap size: 10; running mean: 10; max distance: 300). We then applied the filtering method on the three series with a range of values for the filtering distance (50-500, step size of 50) and identified the value that returned the highest sum of tracks across the three series at the lowest number of mismatches. We present the tracking accuracy before filtering and on the remaining tracks after filtering. We underline that the results presented are thus highly conservative. I.e., the approach could be optimised to extract a higher number of tracks while maintaining the number of mismatches. For example, basing the filtering on the best performing tracking parameters for each series specifically would improve the output of the tracking algorithm. Similarly, the results could be improved by fine-tuning the filtering distance threshold for each series individually. However, as our goal here was to demonstrate that our tracking and filtering method can be applied in a naive setting with fixed values for each parameter we have not included those results.

@@

#
## Tracking and filtering web application

To make the method described here easily accessible for other users, we built a web application through which the trackking and filtering method can be applied on user data. The web application is accessible through https://github.com/TECOLOGYxyz/FlowerTracking.

@@

The results of the tracking parameter test are given in table 1. We present the tracking performance without the use of the three tracking parameters along with the best performance (lowest number of mismatches/highest MOTA), and the best performance where the number of tracks correspond to the number of individual flowers in the series. When several values for a parameter was included in a result, we present the range of values. For two series, 2018 THUL B and 2019 NARS D, the tracking algorithm returned perfect results for several combinations of parameters. For the three remaining series, the algorithm returned consistently high MOTA scores (minimum 0.95). In all cases, except 2019 NARS D, where the tracking algorithm returned perfect results with all parameters zeroed, did the inclusion of the user set tracking parameters improve the tracking accuracy.

In one case, 2019 NARS C, four different numbers of tracks were returned among the parameter combinations that resulted in the lowest number of mismatches. A number of tracks equal to the number of flowers in the series were among these four. In anotehr case, 2019 NYAA E, the parameter combination that produced the highest accuracy did not return a number of tracks equal to the number of flowers in the series. However, combinations that did produce this number of tracks also returned a higher number of mismatches.

**Table 1:** Tracking performance for the automatic flower tracking algorithm for different tracking parameter combinations. Number of tracks in bold shows when the number of tracks equals the number of flowers in the series.
![](../figures/TrackingScoresv22.png){ width=100% }

@@

The value for filtering distance that returned the highest number of tracks at the lowest number of mismatches was 350 (results for all values given in supp. table S1). The results of the filtering algorithm on the three series in which our tracking algorithm did not return perfect results are given in table 2 and the filtering process is visualised in fig. 4. Our filtering method extracted 28 tracks from the three series (2018 NARS A: 15 tracks, 1 mismatch; 2019 NARS C: 10 tracks, 0 mismatches; 2019 NYAA E: 3 tracks, 0 mismatches) with complex scenes using a fixed filtering distance value of 350 and fixed values for the tracking parameters (10, 10, 300, for running mean, max gap size, and max distance, respectively) with only a single mismatch included in the 28 tracks.

**Table 2:** Results of the filtering algorithm with a filtering distance value of 350 applied across all three series.

![](../figures/FilterScoresv211.png){ width=100% }

@@

**Figure 4:** Track filtering process. Row 1, 2, and 3 is 2019 NARS C, 2018 NARS A, and 2019 NYAA E, respectively. First column shows the centroid points in the given series, coloured by the track id from the centroid tracking algorithm. Second column shows the polygons calculated from the tracks. The filtering algorithm with a distance threshold of 350 was applied to the centroids of these polygons. Third column shows the results of the filtering where all tracks that did not pass the filtering have been removed.

# Discussion  
## Tracking

Many tracking algorithms, such as the Simple online and real-time tracking (SORT) [@sort], use Kalman filtering to predict the future location of an object by estimating its velocity [@kalmann]. However, Kalman filtering assumes linear motion of the object and does not handle abrupt motion well. As the flowers are constrained in movement by their stalk, their movement does not follow the assumption of linear motion. Some tracking algorithms, such as DeepSORT [@deepsort2017], take the tracking-by-detection approach in which convolutional neural networks are applied to leverage appearance information in order to continuously distinguish individuals through time, thus aiding the tracking algorithm. In our case, however, flowers are not clearly distinguishable and one flower may look less alike between two frames than two flowers in a single frame. Finally, some tracking methods require manual initiation of a fixed number of tracks for objects in the first frame and the algorithm will track these objects in the subsequent frames [@hu2012]. These methods does not allow new objects to enter the frame through the series. In our case, objects can disappear permanently from the frame, for example when flowers wilt, in which case the track should be terminated. Oppositely, new objects can continuously appear through the series when new flowers bloom of previously unseen flowers enter the frame. Additionally, objects may temporarily disappear from view and may or may not reappear at some later stage. These settings vastly complicate the task of tracking multiple objects. Our tracking algorithm returned consistently accuracy despite the difficult settings.

@@

In a naive setting, a priori knowledge about which settings to use is not available. As with any tracking method, accuracy can only be assessed based on visual observation or more thoroughly by testing on subset of tracks for which a ground truth exists. When applying the method in a new setting, we recommend performing a parameter test as demonstrated here on a subset of ground truth tracks to derive the optimal set of parameters.

@@

The two series for which the tracking algorithm produced perfect results were also the series with the lowest number of flowers. As the number of flowers increase so does the complexity of the task of tracking. A higher number of flowers within the frame increases the possibility of flowers in close vicinity to each other and of flowers overlapping and moving in and out of frame translating to a higher risk of making tracking errors. The very high complexity in the scenes with many flowers, especially 2019 NARS C with 85 flowers, renders it unlikely that any tracking method could produce perfect results. Although this setting can be considered an extreme case for our application, it underlines the relevance of a pipeline including both tracking and filtering.

@@

We presented a filtering method based on the distance between track geometry centroids which can be applied subsequent of tracking. The filtering algorithm successfully extracted correctly tracked flowers from each of the three series for which the tracking algorithm did not produce perfect results. Only a single mismatch was included in the 28 flowers that passed the filtering. Extracting tracks that are spatially isolated does not guarantee that the tracks are correct/without errors. However, as spatially isolated objects are easier to track, it increases confidence in the remaining tracks.
